---
title: "Phoenix Sepsis Score"
output:
 rmarkdown::html_vignette:
   toc: true
   toc_depth: 2
   number_sections: false
bibliography: references.bib
vignette: >
 %\VignetteIndexEntry{Phoenix Sepsis Score}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r label = "setup", include = FALSE}
library(qwraps2)
options(qwraps2_markup = "markdown")
knitr::opts_chunk$set(collapse = TRUE)
```



```{r}
library(phoenix)
packageVersion("phoenix")
```


The of the diagnostic Phoenix Sepsis Criteria developed in
[@sanchez_2024_development](https://doi.org/10.1001/jama.2024.0196) and
[@schlapback_2024_international](https://doi.org/10.1001/jama.2024.0179) is
implimented in this package.

# The Phoenix Sepsis Criteria

The diagnostic Phoenix Sepsis Criteria is defined by dysfuntion in four organ
systems:

|        | 0 Points | 1 Point | 2 Points | 3 Points |
|:-------|:-------- |:------- |:-------- |:-------- |
|**Respiratory** (0-3 points) | PaO<sub>2</sub>:FiO<sub>2</sub> &geq; 400 and SpO<sub>2</sub>:FiO<sub>2</sub> &geq; 292 | (PaO<sub>2</sub>:FiO<sub>2</sub> &lt; 400 or SpO<sub>2</sub>:FiO<sub>2</sub> &lt; 292) and any respiratory support | (PaO<sub>2</sub>:FiO<sub>2</sub> &lt; 200 or SpO<sub>2</sub>:FiO<sub>2</sub> &lt; 220) and IMV | (PaO<sub>2</sub>:FiO<sub>2</sub> &lt; 100 or SpO<sub>2</sub>:FiO<sub>2</sub> &lt; 148) and IMV |
|**Cardiovascular** (0-6 points) |
|&nbsp;&nbsp; _Vasoactive Medications_ | No medications | 1 medictation | 2 or more medictations |
|&nbsp;&nbsp; _Lactate_ | [0, 5) mmol/L | [5, 11) mmol/L | [11, &infin;) mmol/L |
|&nbsp;&nbsp; _MAP_ | | | |
|&nbsp;&nbsp;&nbsp;&nbsp;   0 &leq; Age <   1 | [31, &infin;) mmHg | [17, 31) mmHg | [0, 17) mmHg |
|&nbsp;&nbsp;&nbsp;&nbsp;   1 &leq; Age <  12 | [39, &infin;) mmHg | [25, 39) mmHg | [0, 25) mmHg |
|&nbsp;&nbsp;&nbsp;&nbsp;  12 &leq; Age <  24 | [44, &infin;) mmHg | [31, 44) mmHg | [0, 31) mmHg |
|&nbsp;&nbsp;&nbsp;&nbsp;  24 &leq; Age <  60 | [45, &infin;) mmHg | [32, 45) mmHg | [0, 32) mmHg |
|&nbsp;&nbsp;&nbsp;&nbsp;  60 &leq; Age < 144 | [49, &infin;) mmHg | [36, 49) mmHg | [0, 36) mmHg |
|&nbsp;&nbsp;&nbsp;&nbsp; 144 &leq; Age       | [52, &infin;) mmHg | [38, 52) mmHg | [0, 38) mmHg |
|**Coagulation** (0-2 points) | | 1 point each; max 2 |
|&nbsp;&nbsp; Platelets  | [100, &infin;) K/&mu;L | [0, 100) K/&mu;L |
|&nbsp;&nbsp; INR        | [0 , 1.3] | [1.3, &infin;) |
|&nbsp;&nbsp; D-Dimer    | [0, 2] mg/L FEU | (2, &infin;) mg/L FEU |
|&nbsp;&nbsp; Fibrinogen | [100, &infin;) mg/dL | [0, 100) mg/dL |
|**Neurologic** (0-2 points) | GCS &#8712; {11, 12, 13, 14, 15} | GCS &#8712; {3, 4, ..., 10} | Bilarterally fixed pupils |
<small>
* IMV: invasive mechanical ventalation</br>
* Age: measured in months</br>
* MAP: mean arterial pressure</br>
</small>

A total score of 2 or more points is needed to diagnosis pediatric sepsis.

# The Phoenix 8 Criteria

This extended criteria uses the four organ systems above along with four additional organ dysfunction scores.


|        | 0 Points | 1 Point |
|:-------|:-------- |:------- |
|**Endocrine** (0-1 point) | blood glucose &#8712; [50, 150] mg/dL | blood glucose &#8713; [50, 150] mg/dL |
|**Immunologic** (0-1 point) | ANC &geq; 500 and ALC &geq; 1000 cells /mm<sup>3</sup> | ANC &lt; 500 and/or ALC &lt; 1000 cells/mm<sup>3</sup> |
|**Renal** (0-1 point) |
|&nbsp;&nbsp;   0 &leq; Age <   1 | Creatinine &lt; 0.8 mg/dL | Creatinine &geq; 0.8 mg/dL |
|&nbsp;&nbsp;   1 &leq; Age <  12 | Creatinine &lt; 0.3 mg/dL | Creatinine &geq; 0.3 mg/dL |
|&nbsp;&nbsp;  12 &leq; Age <  24 | Creatinine &lt; 0.4 mg/dL | Creatinine &geq; 0.4 mg/dL |
|&nbsp;&nbsp;  24 &leq; Age <  60 | Creatinine &lt; 0.6 mg/dL | Creatinine &geq; 0.6 mg/dL |
|&nbsp;&nbsp;  60 &leq; Age < 144 | Creatinine &lt; 0.7 mg/dL | Creatinine &geq; 0.7 mg/dL |
|&nbsp;&nbsp; 144 &leq; Age       | Creatinine &lt; 1.0 mg/dL | Creatinine &geq; 1.0 mg/dL |
|**Hepatic** (0-1 point) | Total bilirubin &lt; 4 mg/dL and ALT &gt; 102 IU/L | Total bilirubin %geq; 4 mg/dL and/or ALT &gt; 102 IU/L |
<small>
* Age: measured in months</br>
</small>

# Development of the Criteria

Details on the developed of the criteria are discribed in
@sanchez_2024_development and @schlapback_2024_international and end users
are encouraged to review these papers.  A couple quick notes about the data
and use in general.  Some specific deatils will be provided in each of the
sections for each organ system.

*Missing data = 0 points*: During the development of the Phoenix criteria
missing data was mapped to zero points. This was done as it was reasonable to
assume that for some labs and metrics, missing data could indicate that there
was no concern and the testing was not order.  Further, the Phoenix criteria
was developed to be useful in both high, medium, and low resource settings
where some labs and values migth be uncommon or impossible to get.  As such,
we encourage end users of this package to do the same - missing values are
missing and should not be imputed.

*Worst in first 24 hours*: The score was developed on the worse measured
value during the first 24 hours of an hospital encounter.  But this was also
done on a variable by variable basis.  For example, say one 54 month old
patient had the following cardiovascular scores:

| time from admission | vasoactive medications | lacate  | MAP     |
| :--------:          | :-----:                | :-----: | :-----: |
| 0                   |                        |         | 55      |
| 20                  |                        |         | 55      |
| 120                 |                        |         | 55      |
| 382                 |                        |         | 55      |

# Package Use

There are functions to apply the Phoenix criteria to a data set for each of
the organ systems and wrapers for full Phoenix and Phoenix-8 scores.

Inputs for the organ dysfunction scoring functions are exptected to be
numeric vectors and to have a common length _or_ a length of 1.  For example,
the respiratory score requires four inputs.  The lengths of the inputs could
be
`r  backtick(c(1, 10, 10, 10))  `
indicating that the first input is constant for all other values.  An input
of
`r  backtick(c(2, 10, 10, 10))  `
will result in an error. If you wish to 'recycle' an input of lenght greate
than 1 and less than the max length of another, then wrap the shorter inputs
in
`r  backtick(rep())  `
for clarity.

## Respiratory

### Inputs:
*
`r  backtick(pf_ratio)  `
is the ratio of PaO<sub>2</sub> (partial pressure or oxygen in arterial
blood, units of mmHg) to FiO<sub>2</sub> (fraction of inspiratory oxygen,
values expected to be between 0.21 for room air, to 1.00).  Gathering the
PaO<sub>2</sub> is an invasive procedure.

*
`r  backtick(sf_ratio)  `
The SpO<sub>2</sub> (pulse
oximetry) to FiO<sub>2</sub> ratio is a non-invasive surrogate for the PF
ratio.  Important note: during the development of the Phoenix criteria SF
ratios were only valid to consider if the SpO<sub>2</sub> is &leq; 97.

*
`r  backtick(imv)  `
Invasive mechanical ventalation.  This is an integer valued indicator
variable: 0 = not intubated; 1 = intubated.

*
`r  backtick(other_respiratory_support)  `
Any oxygen support, e.g., high-flow, non-invasive positive pressure, or IMV.

### Scores


```{r echo = FALSE, results = "hide", fig.width = 7, fig.height = 4}
resp_data <-
  expand.grid(pfr = seq(0, 450, by = 10), sfr = seq(0, 450, by = 10), imv = c(0, 1), o2 = c(0, 1))
resp_data$o2 <- pmax(resp_data$imv, resp_data$o2)
resp_data <- unique(resp_data)
resp_data$phoenix_respiratory_score <- factor(phoenix_respiratory(pfr, sfr, imv, o2, data = resp_data))

resp_data$oxygen_support <-
  interaction(resp_data$imv, resp_data$o2) |>
  factor(levels = c("0.0", "0.1", "1.1"), c("No Oxygen Support", "Non-invasive Oxygen Support", "Invasive Oxygen Support"))

ggplot2::ggplot(data = resp_data) +
  ggplot2::theme_classic() +
  ggplot2::aes(x = sfr, y = pfr, fill = phoenix_respiratory_score) +
  ggplot2::geom_tile() +
  ggplot2::facet_wrap( ~ oxygen_support, nrow = 1) +
  ggplot2::scale_fill_brewer(name = "Phoenix Respiratory Score", palette = "Spectral", direction = -1) +
  ggplot2::ylab(expression(PaO[2] : FiO[2])) +
  ggplot2::xlab(expression(SpO[2] : FiO[2])) +
  ggplot2::theme(legend.position = "bottom")
```


### Use

The function
`r  backtick(phoenix_respiratory)  `
will retur

A patient with a nasal cannula at 5 L/min (approximate FiO<sub>2</sub> of
0.4) and SpO<sub>2</sub> of 87:

```{r}
phoenix_respiratory(
  pf_ratio = NA,
  sf_ratio = 87 / 0.40,
  imv = 0,
  other_respiratory_support = 1)
```


When your data is in a
`r  backtick(data.frame)  `
then:

```{r}
DF <- read.table(sep = "|", header = TRUE, text =
"
pfr | sfr | imv | o2
    | 438 |     |
    | 175 |     | 1
    | 175 |     | 1
186 |     |  1  | 0
300 | 277 |  0  | 1
    |     |     |
")

DF$resp_score <- phoenix_respiratory(pfr, sfr, imv, o2, DF)

DF
```


## Cardiovascular

_this section is to be written_

## Coagulation

_this section is to be written_

## Neurologic

_this section is to be written_

## Endocrine

_this section is to be written_

## Immunologic

_this section is to be written_

## Renal

_this section is to be written_

## Hepatic

_this section is to be written_

## Phoenix

_this section is to be written_

## Phonix 8

_this section is to be written_

# References
<div id="refs"></div>

# Session Info

```{r label = "sessioninfo"}
sessionInfo()
```

