---
title: "The Phoenix Sepsis Criteria in python"
output:
 rmarkdown::html_vignette:
   toc: true
   toc_depth: 2
   number_sections: false
bibliography: references.bib
vignette: >
 %\VignetteIndexEntry{The Phoenix Sepsis Criteria in python}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, eval = FALSE, include = FALSE}
################################################################################
#                                IMPORTANT NOTE                                #
#                                                                              #
#   This vignette _is not_ part of the R pacakge, it is part of the website!   #
################################################################################
```

```{css, echo = FALSE}
.rchunk {
  border: 5px solid #FFFFFF;
}
.pychunk {
  border: 5px solid #4B8BBE;
}
```

```{r, label = setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(reticulate)

if (!virtualenv_exists(envname = "phoenix")) {
  virtualenv_create(envname = "phoenix")
  virtualenv_install(envname = "phoenix",
                     packages = c("numpy", "pyarrow", "pandas"))
}

use_virtualenv(virtualenv = "phoenix")
```


For those who are primarily working in python and want or need to impliment the
Phoenix Sepsis criteria there are a couple options.

1. Use [rpy2](https://rpy2.github.io/index.html) to use the phoenix R package
   within python

2. Use the this rudimentary [python module](https://github.com/dewittpe/phoenix/tree/main/python)
   Within that directory you will find the module, example data set, and
   example script.  Most of that is reproduced in this article.

For the following examples you'll need the following python modules
```{python, include = FALSE}
import sys
sys.path.append('../python')
```

```{python class.source="pychunk"}
import numpy as np
import pandas as pd
import phoenix as phx
```

Start by reading in the example data set.  This is the same data as provided in
the R pacakge.
```{python, include = FALSE}
# use this block to get read the data, use an unevaluated chunk to make it look
# good in the article
sepsis = pd.read_csv("../python/sepsis.csv")
```

```{python class.source = "pychunk", eval = FALSE}
sepsis = pd.read_csv("sepsis.csv")
```

We will compare the python output to the R results too.
```{r class.source = "rchunk"}
library(phoenix)
```
(In case you didn't notice, python code chunks are outlined in blue to
distinguish them from R code chunks.)

# Organ Dysfunction Scores

## Respiratory

For the respiratory score, as with the other organ system dysfunction scores,
the return in R is an integer vector.  In python you should expect a similar
return of a numpy integer array.

In R, we got the respiratory score via:
```{r class.source = "rchunk"}
r_resp = phoenix_respiratory(
  pf_ratio = pao2 / fio2,
  sf_ratio = ifelse(spo2 <= 97, spo2 / fio2, NA_real_),
  imv = vent,
  other_respiratory_support = as.integer(fio2 > 0.21),
  data = sepsis
  )
```
In python you have a similar syntax
```{python class.source = "pychunk", output.var = "python_respiratory"}
py_resp = phx.phoenix_respiratory(
    pf_ratio = sepsis["pao2"] / sepsis["fio2"],
    sf_ratio = np.where(sepsis["spo2"] <= 97, sepsis["spo2"] / sepsis["fio2"], np.nan),
    imv      = sepsis["vent"],
    other_respiratory_support = (sepsis["fio2"] > 0.21).astype(int).to_numpy()
)
```

```{r, include = FALSE}
stopifnot(identical(r_resp, as.integer(py$py_resp)))
```
```{r}
identical(r_resp, as.integer(py$py_resp))
```


## Cardiovascular

```{r class.source = 'rchunk'}
r_card <- phoenix_cardiovascular(
  vasoactives = dobutamine + dopamine + epinephrine + milrinone + norepinephrine + vasopressin,
  lactate = lactate,
  age = age,
  map = dbp + (sbp - dbp)/3,
  data = sepsis
)
```

```{python class.source = "pychunk"}
py_card = phx.phoenix_cardiovascular(
    vasoactives = sepsis["dobutamine"] + sepsis["dopamine"] + sepsis["epinephrine"] + sepsis["milrinone"] + sepsis["norepinephrine"] + sepsis["vasopressin"],
    lactate = sepsis["lactate"],
    age = sepsis["age"],
    map = sepsis["dbp"] + (sepsis["sbp"] - sepsis["dbp"]) / 3
)
```

```{r, include = FALSE}
stopifnot(identical(r_card, as.integer(py$py_card)))
```
```{r}
identical(r_card, as.integer(py$py_card))
```

## Coagulation

```{r class.source = 'rchunk'}
r_coag <-
  phoenix_coagulation(platelets, inr, d_dimer, fibrinogen, data = sepsis)
```

```{python class.source = "pychunk"}
py_coag = phx.phoenix_coagulation(
    platelets = sepsis['platelets'],
    inr = sepsis['inr'],
    d_dimer = sepsis['d_dimer'],
    fibrinogen = sepsis['fibrinogen']
)
```

```{r, include = FALSE}
stopifnot(identical(r_coag, as.integer(py$py_coag)))
```
```{r}
identical(r_coag, as.integer(py$py_coag))
```

## Neurologic

```{r class.source = 'rchunk'}
r_neur <-
  phoenix_neurologic(
    gcs = gcs_total,
    fixed_pupils = as.integer(pupil == "both-fixed"),
    data = sepsis
  )
```

```{python class.source = "pychunk"}
py_neur = phx.phoenix_neurologic(
    gcs = sepsis["gcs_total"],
    fixed_pupils = (sepsis["pupil"] == "both-fixed").astype(int)
)
```

```{r, include = FALSE}
r_neur
py$py_neur
stopifnot(identical(r_neur, as.integer(py$py_neur)))
```
```{r}
identical(r_neur, as.integer(py$py_neur))
```

## Endocrine

```{r class.source = 'rchunk'}
r_endo <- phoenix_endocrine(glucose, data = sepsis)
```

```{python class.source = "pychunk"}
py_endo = phx.phoenix_endocrine(sepsis["glucose"])
```

```{r, include = FALSE}
stopifnot(identical(r_endo, as.integer(py$py_endo)))
```
```{r}
identical(r_endo, as.integer(py$py_endo))
```

## Immunologic

```{r class.source = 'rchunk'}
r_immu <- phoenix_immunologic(anc, alc, sepsis)
```

```{python class.source = "pychunk"}
py_immu = phx.phoenix_immunologic(sepsis["anc"], sepsis["alc"])
```

```{r, include = FALSE}
stopifnot(identical(r_immu, as.integer(py$py_immu)))
```
```{r}
identical(r_immu, as.integer(py$py_immu))
```

## Renal

```{r class.source = 'rchunk'}
r_renal <- phoenix_renal(creatinine = creatinine, age = age, data = sepsis)
```

```{python class.source = "pychunk"}
py_renal = phx.phoenix_renal(sepsis["creatinine"], sepsis["age"])
```

```{r, include = FALSE}
stopifnot(identical(r_renal, as.integer(py$py_renal)))
```
```{r}
identical(r_renal, as.integer(py$py_renal))
```

## Hepatic

```{r class.source = 'rchunk'}
r_hepatic <- phoenix_hepatic(bilirubin = bilirubin, alt = alt, data = sepsis)
```

```{python class.source = "pychunk"}
py_hepatic = phx.phoenix_hepatic(sepsis["bilirubin"], sepsis["alt"])
```

```{r, include = FALSE}
stopifnot(identical(r_hepatic, as.integer(py$py_hepatic)))
```
```{r}
identical(r_hepatic, as.integer(py$py_hepatic))
```


# Phoenix

In R, the call to `phoenix` will return a `data.frame`.  In python the return is
a pandas DataFrame.

```{r class.source = 'rchunk'}
r_phoenix_scores <-
  phoenix(
    # respiratory
      pf_ratio = pao2 / fio2,
      sf_ratio = ifelse(spo2 <= 97, spo2 / fio2, NA_real_),
      imv = vent,
      other_respiratory_support = as.integer(fio2 > 0.21),
    # cardiovascular
      vasoactives = dobutamine + dopamine + epinephrine + milrinone + norepinephrine + vasopressin,
      lactate = lactate,
      age = age,
      map = dbp + (sbp - dbp)/3,
    # coagulation
      platelets = platelets,
      inr = inr,
      d_dimer = d_dimer,
      fibrinogen = fibrinogen,
    # neurologic
      gcs = gcs_total,
      fixed_pupils = as.integer(pupil == "both-fixed"),
    data = sepsis
  )
```

```{python class.source = "pychunk"}
py_phoenix_scores = phx.phoenix(
    pf_ratio = sepsis["pao2"] / sepsis["fio2"],
    sf_ratio = np.where(sepsis["spo2"] <= 97, sepsis["spo2"] / sepsis["fio2"], np.nan),
    imv      = sepsis["vent"],
    other_respiratory_support = (sepsis["fio2"] > 0.21).astype(int).to_numpy(),
    vasoactives = sepsis["dobutamine"] + sepsis["dopamine"] + sepsis["epinephrine"] + sepsis["milrinone"] + sepsis["norepinephrine"] + sepsis["vasopressin"],
    lactate = sepsis["lactate"],
    age = sepsis["age"],
    map = sepsis["dbp"] + (sepsis["sbp"] - sepsis["dbp"]) / 3,
    platelets = sepsis['platelets'],
    inr = sepsis['inr'],
    d_dimer = sepsis['d_dimer'],
    fibrinogen = sepsis['fibrinogen'],
    gcs = sepsis["gcs_total"],
    fixed_pupils = (sepsis["pupil"] == "both-fixed").astype(int),
    )
```

```{r, include = FALSE}
r_phoenix_scores
py$py_phoenix_scores
stopifnot(
all.equal(r_phoenix_scores, as.data.frame(py$py_phoenix_scores),
          check.attributes = FALSE)
)
```
```{r}
all.equal(r_phoenix_scores, as.data.frame(py$py_phoenix_scores),
          check.attributes = FALSE)
```

# Phoenix8

```{r class.source = 'rchunk'}
r_phoenix8_scores <-
  phoenix8(
    # respiratory
      pf_ratio = pao2 / fio2,
      sf_ratio = ifelse(spo2 <= 97, spo2 / fio2, NA_real_),
      imv = vent,
      other_respiratory_support = as.integer(fio2 > 0.21),
    # cardiovascular
      vasoactives = dobutamine + dopamine + epinephrine + milrinone + norepinephrine + vasopressin,
      lactate = lactate,
      age = age, # Also used in the renal assessment.
      map = dbp + (sbp - dbp)/3,
    # coagulation
      platelets = platelets,
      inr = inr,
      d_dimer = d_dimer,
      fibrinogen = fibrinogen,
    # neurologic
      gcs = gcs_total,
      fixed_pupils = as.integer(pupil == "both-fixed"),
    # endocrine
      glucose = glucose,
    # immunologic
      anc = anc,
      alc = alc,
    # renal
      creatinine = creatinine,
      # no need to specify age again
    # hepatic
      bilirubin = bilirubin,
      alt = alt,
    data = sepsis
  )
```

```{python class.source = "pychunk"}
py_phoenix8_scores = phx.phoenix8(
    pf_ratio = sepsis["pao2"] / sepsis["fio2"],
    sf_ratio = np.where(sepsis["spo2"] <= 97, sepsis["spo2"] / sepsis["fio2"], np.nan),
    imv      = sepsis["vent"],
    other_respiratory_support = (sepsis["fio2"] > 0.21).astype(int).to_numpy(),
    vasoactives = sepsis["dobutamine"] + sepsis["dopamine"] + sepsis["epinephrine"] + sepsis["milrinone"] + sepsis["norepinephrine"] + sepsis["vasopressin"],
    lactate = sepsis["lactate"],
    map = sepsis["dbp"] + (sepsis["sbp"] - sepsis["dbp"]) / 3,
    platelets = sepsis['platelets'],
    inr = sepsis['inr'],
    d_dimer = sepsis['d_dimer'],
    fibrinogen = sepsis['fibrinogen'],
    gcs = sepsis["gcs_total"],
    fixed_pupils = (sepsis["pupil"] == "both-fixed").astype(int),
    glucose = sepsis["glucose"],
    anc = sepsis["anc"],
    alc = sepsis["alc"],
    creatinine = sepsis["creatinine"],
    bilirubin = sepsis["bilirubin"],
    alt = sepsis["alt"],
    age = sepsis["age"]
    )
```

```{r, include = FALSE}
stopifnot(
  all.equal(r_phoenix8_scores, as.data.frame(py$py_phoenix8_scores),
            check.attributes = FALSE)
)
```
```{r}
all.equal(r_phoenix8_scores, as.data.frame(py$py_phoenix8_scores),
          check.attributes = FALSE)
```

