---
title: "The Phoenix Sepsis Criteria in sql"
output:
 rmarkdown::html_vignette:
   toc: true
   toc_depth: 2
   number_sections: false
bibliography: references.bib
vignette: >
 %\VignetteIndexEntry{The Phoenix Sepsis Criteria in sql}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, eval = FALSE, include = FALSE}
################################################################################
#                                IMPORTANT NOTE                                #
#                                                                              #
#   This vignette _is not_ part of the R package, it is part of the website!   #
################################################################################
```

```{css, echo = FALSE}
.rchunk {
  border: 5px solid #FFFFFF;
}
.sqlchunk {
  border: 5px solid #4B8BBE;
}
```

```{r, label = setup, include = FALSE}
for (pkg in c("odbc", "DBI", "RSQLite")) {
  if (!(pkg %in% rownames(installed.packages()))) {
    install.packages(pkg, repo = "https://cran.rstudio.com")
  }
}

knitr::opts_chunk$set(collapse = TRUE)
```

The following examples are provided to show how to apply the Phoenix Sepsis
criteria within an SQLite data base.  We will use the same example data set
`sepsis` from the R package for the following examples.  The examples
are made possible by using the following R packages
```{r}
library(odbc)
library(DBI)
library(RSQLite)
library(phoenix)
```

To help distinguish between R and sql code chunks, chunks outlined in blue are
SQL and chunks with no outline are R chunks.

We start in R by setting up a local example SQL database with one table `sepsis`

```{r}
con <- dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")
dbWriteTable(conn = con, name = "sepsis", value = sepsis)
```

We can easily query the table:

```{sql class.source = 'sqlchunk', connection = con}
SELECT * FROM sepsis
```

# Respiratory

We need either create or code for the PF ratio, SF Ratio, invasive mechanical
ventilation, and other respiratory support.  In the construction of the PF ratio
(pfr) and SF ratio (sfr) we replace missing values with value that will map to a
score of zero.  If the percent of inspiratory oxygen is greater than 0.21, that
indicates the patient is on some form of respiratory support.

```{sql class.source = 'sqlchunk', connection = con, output.var = "resp"}
WITH respiratory AS (
  SELECT
    pid,
    COALESCE(pao2 / fio2, 500) AS pfr,
    COALESCE(spo2 / fio2, 500) AS sfr,
    COALESCE(vent, 0)          AS imv,
    IIF(fio2 > 0.21 OR vent = 1, 1, 0) AS other_respiratory_support
  FROM sepsis
)
SELECT
  pid,
  imv * (IIF(pfr < 100 OR sfr < 14, 1, 0) + IIF(pfr < 200 OR sfr < 220, 1, 0) ) +
    other_respiratory_support * IIF(pfr < 400 OR sfr < 292, 1, 0) AS phoenix_respiratory_score
  FROM respiratory;
```

The above query was saved to an R `data.frame` called `resp` and will be added
to the virtual database for inspection later.

```{r}
dbWriteTable(conn = con, name = "respiratory", value = resp)
```

# Cardiovascular

Missing values correspond to zero points.  The combination of age and mean
arterial pressure will require both values to be know for possible points to be
assessed.

```{sql class.source = 'sqlchunk', connection = con, output.var = "card"}
WITH vars AS (
  SELECT
    pid,
    COALESCE(dobutamine, 0) +
      COALESCE(dopamine, 0) +
      COALESCE(epinephrine, 0) +
      COALESCE(milrinone, 0) +
      COALESCE(norepinephrine, 0) +
      COALESCE(vasopressin, 0) AS vasos,
    lactate,
    dbp + (1/3) * (sbp - dbp) AS map,
    age
  FROM sepsis
),
points AS (
  SELECT
    pid,
    CASE WHEN vasos > 1 THEN 2
         WHEN vasos > 0 THEN 1
         ELSE 0 END AS vaso_points,
    CASE WHEN lactate >= 11 THEN 2
         WHEN lactate >=  5 THEN 1
         ELSE 0 END AS lactate_points,
    CASE WHEN (               age <   1) AND (map < 17) THEN 2
         WHEN (age >=   1 AND age <  12) AND (map < 25) THEN 2
         WHEN (age >=  12 AND age <  24) AND (map < 31) THEN 2
         WHEN (age >=  24 AND age <  60) AND (map < 32) THEN 2
         WHEN (age >=  60 AND age < 144) AND (map < 36) THEN 2
         WHEN (age >= 144              ) AND (map < 38) THEN 2
         WHEN (               age <   1) AND (map < 31) THEN 1
         WHEN (age >=   1 AND age <  12) AND (map < 39) THEN 1
         WHEN (age >=  12 AND age <  24) AND (map < 44) THEN 1
         WHEN (age >=  24 AND age <  60) AND (map < 45) THEN 1
         WHEN (age >=  60 AND age < 144) AND (map < 49) THEN 1
         WHEN (age >= 144              ) AND (map < 52) THEN 1
         ELSE 0 END AS map_points
  FROM vars
)
SELECT
  pid,
  vaso_points + lactate_points + map_points AS phoenix_cardiovascular_score
FROM points;
```

The results of the above query were saved behind the scene to an R `data.frame`
called `card` which will be added to the SQLite database for inspection later.

```{r}
dbWriteTable(conn = con, name = "cardiovascular", value = card)
```

# Coagulation

```{sql class.source = 'sqlchunk', connection = con, output.var = "coag"}
WITH points AS (
  SELECT
    pid,
    CASE WHEN platelets < 100 THEN 1 ELSE 0 END AS plts,
    CASE WHEN inr > 1.3 THEN 1 ELSE 0 END AS inr,
    CASE WHEN d_dimer > 2 THEN 1 ELSE 0 END AS ddm,
    CASE WHEN fibrinogen < 100 THEN 1 ELSE 0 END AS fib
  FROM sepsis
)
SELECT
  pid,
  CASE WHEN plts + inr + ddm + fib >= 2 THEN 2
       ELSE plts + inr + ddm + fib END AS phoenix_coagulation_score
  FROM points
```

The above was stored as an R `data.frame` named `coag` and will be added to the
SQLite database for inspection later.

```{r}
dbWriteTable(conn = con, name = "coagulation", value = coag)
```

# Neurologic

```{sql class.source = 'sqlchunk', connection = con, output.var = "neuro"}
WITH points AS (
  SELECT
    pid,
    CASE WHEN gcs_total <= 10 THEN 1 ELSE 0 END AS gcs,
    CASE WHEN pupil = "both-fixed" THEN 1 ELSE 0 END AS fixed_pupils
  FROM sepsis
)
SELECT
  pid,
  CASE WHEN fixed_pupils = 1 THEN 2
       WHEN gcs = 1 THEN 1
       ELSE 0 END as phoenix_neurologic_score
  FROM points
```

The above was stored as an R `data.frame` named `neuro` and will be added to the
SQLite database for inspection later.

```{r}
dbWriteTable(conn = con, name = "neurologic", value = neuro)
```

# Endocrine
```{sql class.source = 'sqlchunk', connection = con, output.var = "endo"}
SELECT
  pid,
  CASE WHEN glucose <  50 THEN 1
       WHEN glucose > 150 THEN 1
       ELSE 0 END AS phoenix_endocrine_score
  FROM sepsis
```

The above was stored as an R `data.frame` named `endo` and will be added to the
SQLite database for inspection later.

```{r}
dbWriteTable(conn = con, name = "endocrine", value = endo)
```

# Immunologic
```{sql class.source = 'sqlchunk', connection = con, output.var = "immu"}
SELECT
  pid,
  CASE WHEN anc <  500 THEN 1
       WHEN alc < 1000 THEN 1
       ELSE 0 END AS phoenix_immunologic_score
FROM sepsis
```

The above was stored as an R `data.frame` named `immu` and will be added to the
SQLite database for inspection later.
```{r}
dbWriteTable(conn = con, name = "immunologic", value = immu)
```

# Renal
```{sql class.source = 'sqlchunk', connection = con, output.var = "renal"}
SELECT
  pid,
  CASE WHEN (               age <   1) AND (creatinine < 0.8) THEN 1
       WHEN (age >=   1 AND age <  12) AND (creatinine < 0.3) THEN 1
       WHEN (age >=  12 AND age <  24) AND (creatinine < 0.4) THEN 1
       WHEN (age >=  24 AND age <  60) AND (creatinine < 0.6) THEN 1
       WHEN (age >=  60 AND age < 144) AND (creatinine < 0.7) THEN 1
       WHEN (age >= 144              ) AND (creatinine < 1.0) THEN 1
       ELSE 0 END AS phoenix_renal_score
FROM sepsis
```

The above was stored as an R `data.frame` named `renal` and will be added to the
SQLite database for inspection later.
```{r}
dbWriteTable(conn = con, name = "renal", value = renal)
```

# Hepatic
```{sql class.source = 'sqlchunk', connection = con, output.var = "hep"}
SELECT
  pid,
  CASE WHEN bilirubin >= 4 THEN 1
       WHEN alt < 102 THEN 1
       ELSE 0 END AS phoenix_hepatic_score
FROM sepsis
```
The above was stored as an R `data.frame` named `hep` and will be added to the
SQLite database for inspection later.
```{r}
dbWriteTable(conn = con, name = "hepatic", value = hep)
```

# Phoenix

In R and python we provided wrappers for the total score.  Here, we will join
together the results and get the totals.  The results of the following query are
saved in an R `data.frame` called `phoenix_scores`.

```{sql class.source = 'sqlchunk', connection = con, output.var = "phoenix_scores"}
SELECT
  respiratory.pid AS pid,
  phoenix_respiratory_score,
  phoenix_cardiovascular_score,
  phoenix_coagulation_score,
  phoenix_neurologic_score,
  phoenix_respiratory_score + phoenix_cardiovascular_score +
    phoenix_coagulation_score + phoenix_neurologic_score AS phoenix_sepsis_score,
  IIF(phoenix_respiratory_score + phoenix_cardiovascular_score +
    phoenix_coagulation_score + phoenix_neurologic_score >=2, 1, 0) AS phoenix_sepsis,
  IIF(phoenix_respiratory_score + phoenix_cardiovascular_score +
    phoenix_coagulation_score + phoenix_neurologic_score >=2 AND phoenix_cardiovascular_score > 0, 1, 0) AS phoenix_septic_shock
FROM respiratory
LEFT JOIN cardiovascular
ON respiratory.pid = cardiovascular.pid
LEFT JOIN coagulation
ON respiratory.pid = coagulation.pid
LEFT JOIN neurologic
ON respiratory.pid = neurologic.pid
```

```{r}
phoenix_scores
```

This output is the same as the output generated in R via:
```{r}
phoenix_scores_from_r <-
  phoenix(
    # respiratory
      pf_ratio = pao2 / fio2,
      sf_ratio = ifelse(spo2 <= 97, spo2 / fio2, NA_real_),
      imv = vent,
      other_respiratory_support = as.integer(fio2 > 0.21),
    # cardiovascular
      vasoactives = dobutamine + dopamine + epinephrine + milrinone + norepinephrine + vasopressin,
      lactate = lactate,
      age = age,
      map = dbp + (sbp - dbp)/3,
    # coagulation
      platelets = platelets,
      inr = inr,
      d_dimer = d_dimer,
      fibrinogen = fibrinogen,
    # neurologic
      gcs = gcs_total,
      fixed_pupils = as.integer(pupil == "both-fixed"),
    data = sepsis
  )
str(phoenix_scores)
str(phoenix_scores_from_r)

identical(phoenix_scores[, -1], phoenix_scores_from_r)
```



# Phoenix 8


```{r}
dbDisconnect(con)
```


