---
title: "The Phoenix Sepsis Criteria in sql"
output:
 rmarkdown::html_vignette:
   toc: true
   toc_depth: 2
   number_sections: false
bibliography: references.bib
vignette: >
 %\VignetteIndexEntry{The Phoenix Sepsis Criteria in sql}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

```{r, eval = FALSE, include = FALSE}
################################################################################
#                                IMPORTANT NOTE                                #
#                                                                              #
#   This vignette _is not_ part of the R package, it is part of the website!   #
################################################################################
```

```{css, echo = FALSE}
.rchunk {
  border: 5px solid #FFFFFF;
}
.sqlchunk {
  border: 5px solid #4B8BBE;
}
```

```{r, label = setup, include = FALSE}
for (pkg in c("odbc", "DBI", "RSQLite")) {
  if (!(pkg %in% rownames(installed.packages()))) {
    install.packages(pkg, repo = "https://cran.rstudio.com")
  }
}

knitr::opts_chunk$set(collapse = TRUE)
```

The following examples are provided to show how to apply the Phoenix Sepsis
criteria within an SQLite data base.  We will use the same example data set
`sepsis` from the R package for the following examples.  The examples
are made possible by using the following R packages
```{r}
library(odbc)
library(DBI)
library(RSQLite)
library(phoenix)
```

To help distinguish between R and sql code chunks, chunks outlined in blue are
SQL and chunks with no outline are R chunks.

We start in R by setting up a local example SQL database with one table `sepsis`

```{r}
con <- dbConnect(drv = RSQLite::SQLite(), dbname = ":memory:")
dbWriteTable(conn = con, name = "sepsis", value = sepsis)
```

We can easily query the table:

```{sql class.source = 'sqlchunk', connection = con}
SELECT * FROM sepsis
```

# Respiratory

We need either create or code for the PF ratio, SF Ratio, invasive mechanical
ventilation, and other respiratory support.  In the construction of the PF ratio
(pfr) and SF ratio (sfr) we replace missing values with value that will map to a
score of zero.  If the percent of inspiratory oxygen is greater than 0.21, that
indicates the patient is on some form of respiratory support.

```{sql class.source = 'sqlchunk', connection = con, output.var = "resp"}
WITH respiratory AS (
  SELECT
    COALESCE(pao2 / fio2, 500) AS pfr,
    COALESCE(spo2 / fio2, 500) AS sfr,
    COALESCE(vent, 0)          AS imv,
    IIF(fio2 > 0.21 OR vent = 1, 1, 0) AS other_respiratory_support
  FROM sepsis
)
SELECT
  imv * (IIF(pfr < 100 OR sfr < 14, 1, 0) + IIF(pfr < 200 OR sfr < 220, 1, 0) ) +
    other_respiratory_support * IIF(pfr < 400 OR sfr < 292, 1, 0) AS phoenix_respiratory_score
  FROM respiratory;
```

The above query was saved to an R `data.frame` called `resp` and will be added
to the virtual database for inspection later.

```{r}
dbWriteTable(conn = con, name = "respiratory", value = resp)
```

# Cardiovascular

Missing values correspond to zero points.  The combination of age and mean
arterial pressure will require both values to be know for possible points to be
assessed.

```{sql class.source = 'sqlchunk', connection = con, output.var = "card"}
WITH vars AS (
  SELECT
    COALESCE(dobutamine, 0) +
      COALESCE(dopamine, 0) +
      COALESCE(epinephrine, 0) +
      COALESCE(milrinone, 0) +
      COALESCE(norepinephrine, 0) +
      COALESCE(vasopressin, 0) AS vasos,
    lactate,
    dbp + (1/3) * (sbp - dbp) AS map,
    age
  FROM sepsis
),
points AS (
  SELECT
    CASE WHEN vasos > 1 THEN 2
         WHEN vasos > 0 THEN 1
         ELSE 0 END AS vaso_points,
    CASE WHEN lactate >= 11 THEN 2
         WHEN lactate >=  5 THEN 1
         ELSE 0 END AS lactate_points,
    CASE WHEN (               age <   1) AND (map < 17) THEN 2
         WHEN (age >=   1 AND age <  12) AND (map < 25) THEN 2
         WHEN (age >=  12 AND age <  24) AND (map < 31) THEN 2
         WHEN (age >=  24 AND age <  60) AND (map < 32) THEN 2
         WHEN (age >=  60 AND age < 144) AND (map < 36) THEN 2
         WHEN (age >= 144              ) AND (map < 38) THEN 2
         WHEN (               age <   1) AND (map < 31) THEN 1
         WHEN (age >=   1 AND age <  12) AND (map < 39) THEN 1
         WHEN (age >=  12 AND age <  24) AND (map < 44) THEN 1
         WHEN (age >=  24 AND age <  60) AND (map < 45) THEN 1
         WHEN (age >=  60 AND age < 144) AND (map < 49) THEN 1
         WHEN (age >= 144              ) AND (map < 52) THEN 1
         ELSE 0 END AS map_points
  FROM vars
)
SELECT
  vaso_points + lactate_points + map_points AS phoenix_cardiovascular_score
FROM points;
```

The results of the above query were saved behind the scene to an R `data.frame`
called `card` which will be added to the SQLite database for inspection later.

```{r}
dbWriteTable(conn = con, name = "cardiovascular", value = card)
```

# Coagulation

# Neurologic

# Endocrine

# Immunologic

# Renal

# Hepatic

# Phoenix

```{sql eval = FALSE, class.source = 'sqlchunk', connection = con}
```

# Phoenix 8


```{r}
dbDisconnect(con)
```


