---
title: "Example Data Preperation and Scoring"
subtitle: "__THIS IS A WORK IN PROGRESS__"
bibliography: ../references.bib
---

In this document we will illustrate the steps, in sequence, used to prepare data
retrospective data for analysis.

Expected general structure of the data.

* There is a patient and/or encounter id
* Some element for timing is available.  In our work we used a variable called
  `eclock` for encounter-clock.  The values of `eclcok` start at 0, the time of
  admission to the facility, and increments by minutes.  The value `eclock = 5`
  is five minutes after admission.  `eclock = 1440` is 24 hours after admission.

* Not all inputs are reported at the same time.
  Last-observation-carried-forward will be used.

For this example we have the following data:


**TODO: Make the example data easy for end users to access**

```{r}
egdata <- data.table::fread("phx.csv")
str(egdata)
```

The data set contains the following variables.  Details on the expected value
ranges and use will follow as will the details on the caveats.

```{r, echo = FALSE, results = "asis"}
inputvars <- data.table::fread("inputvars.md")
kableExtra::kbl(inputvars) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped")
  )
```

# Data Range Checks

These checks are for obviously impossible values.  You may concider more
conservative checks in your own work.

## FiO~2~

All value should be within [0.21, 1.00], ignoring missing values.  There are no
units to check here.
```{r}
stopifnot(
  all(egdata[["fio2"]] >= 0.21, na.rm = TRUE),
  all(egdata[["fio2"]] <= 1.00, na.rm = TRUE)
)
```

## SpO~2~
All values should be within [0, 100].  Generally, we should expect
integer values, but that may not be needed.  Ignore missing values.  We do have
a caveat on the use of SpO~2~ for values > 97 which will be address later in
this work.  There are no units to check here.

```{r}
stopifnot(
  all(egdata[["spo2"]] >=   0, na.rm = TRUE),
  all(egdata[["spo2"]] <= 100, na.rm = TRUE)
)
```

## PaO~2~
All values should be non-negative.  Expected units of measure of mmHg.  Units
are not in the example data set and not shown in this example.

```{r}
stopifnot(
  all(egdata[["pao2"]] >= 0, na.rm = TRUE)
)
```

## Invasive Mechanical Vent
We expect that the variable used to report invasive mechanical ventilation (IMV)
is an indicator variable: 1 if on IMV, 0 otherwise.  We check here for 0s and 1s
and will consider the duration of IMV later.

```{r}
stopifnot(
  all(egdata[["vent"]] >= 0, na.rm = TRUE)
)
```

## Vasoactive Medications
We expect that the six vasoactive medications are reported as indicators for
system use.  For the checks here we expect all the values to be missing, 0, or
1.

```{r, error = TRUE}
stopifnot(
  all(egdata[["dobutamine"]]     %in% c(NA, 0, 1)),
  all(egdata[["dopamine"]]       %in% c(NA, 0, 1)),
  all(egdata[["epinephrine"]]    %in% c(NA, 0, 1)),
  all(egdata[["milrinone"]]      %in% c(NA, 0, 1)),
  all(egdata[["norepinephrine"]] %in% c(NA, 0, 1)),
  all(egdata[["vasopressin"]]    %in% c(NA, 0, 1))
)
```
## dobutamine
## dopamine
## epinephrine
## milrinone
## norepinephrine
## vasopressin
## lactate
## sbp_art
## dbp_art
## map_art
## sbp_cuff
## dbp_cuff
## map_cuff
## platelets
## inr
## fibrinogen
## d_dimer
## gcs_motor
## gcs_verbal
## gcs_eye
## pupil
