---
title: "Phoenix Sepsis EHR Implementation Notes"
subtitle: "__THIS IS A WORK IN PROGRESS__"
bibliography: ../references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Phoenix criteria for pediatric sepsis and septic shock was developed using
an international data set of pediatric hospitalizations.  Note that
birth-hospitalizations and patients with a gestational age less than 37 weeks
were omitted [@sanchezpinto_2024_development;@schlapbach_2024_international].
The Phoenix criteria were developed based on modeling hospital mortality based
on suspected infection and organ dysfunction within the first twenty-four hours
of a hospital encounter.

The purpose of this document is to provide details on the expected inputs for
the scoring.  The phoenix R package and Python module, along with the example
SQL code assumes the data has been prepared as expected and returns scores
accordingly.

<!-- NOTES:
D_dimer mg/l feu <= 500
GCS Total - gcs_total >= 3 and <= 15 (i.e. GCS should be between 3 and 15)
Glucose - Glucose mg/dl >= 5 and Glucose mg/dl <= 2000
-->

# General Data Formatting Notes

* `NULL` values are treated as non-score generating criteria. Do not populate null/unknown/not applicable values with estimates or outdated information.
* Values outside ranges mentioned below are treated as `NULL`
* Before sending data to the package, verify data. Some values may provide warnings, others may throw errors. Verify your processing can handle warnings and errors.
* Ideally input values will be known along with a time of the measurement so that last observation carried forward logic can be implemented with limits and conditions for age of inputs for calculated variables can be verified.

# Suspected Infection

For a patient to have sepsis or septic shock based on the Phoenix criteria, they
must have a suspected infection.  A patient is considered to have a suspected
infection if they have:

1. an infective test _ordered_, and
2. at least two doses of systemic antimicrobial medication(s).

Note that the test, e.g., a blood culture, need only be ordered.
Positive/negative test status is not needed.

Systemic antimicrobial medications are any antibiotic, antiviral, or antifungal
medication which had any of the following routes:

* ASSUMED SYSTEMIC
* ENTERAL
* HEMODIALYSIS
* INTRA-ARTERIAL
* INTRACARDIAC
* INTRADUODENAL
* INTRAGASTRIC
* INTRAILEAL
* INTRAMUSCULAR
* INTRAVASCULAR
* INTRAVENOUS
* INTRAVENOUS BOLUS
* INTRAVENOUS DRIP
* NASOGASTRIC
* ORAL
* OROPHARYNGEAL
* PARENTERAL
* RECTAL
* SUBCUTANEOUS

Example SQL code used for mapping medication routes when developing Phoenix can be found in
a [GitHub archive](https://github.com/CU-DBMI-Peds/phoenix_sepsis_criteria/tree/main/harmonization/medication_mappings)
specifically the files:
[med_route_mapping.sql](https://github.com/CU-DBMI-Peds/phoenix_sepsis_criteria/blob/main/harmonization/medication_mappings/med_route_mapping.sql)
and
[clean_up_routes.sql](https://github.com/CU-DBMI-Peds/phoenix_sepsis_criteria/blob/main/harmonization/medication_mappings/clean_up_routes.sql).

# Respiratory

Respiratory dysfunction scoring is based on PaO~2~/FiO~2~ (PFR), SpO~2~/FiO~2~
(SFR) and respiratory support.  While the scoring is based on SFR and PFR, it is
preferable to get the SpO~2~, PaO~2~, and FiO~2~ values so certain data checks
can be performed before the score is determined.

## Inputs

### PaO~2~: Partial pressure of oxygen in arterial blood
* Units: mmHg
* Values: [0, Inf)
* Caveats:
* Look Back: 6 hours

### SpO~2~: Peripheral oxygen saturation (pulse ox)
* Units: percentage
* Values: [0, 100]
  * Expect integer values 0, 1, 2, ..., 100
* Caveats: values greater than 97 should _not_ be used for SFR
* Look back: 6 hours

### FiO~2~: Percent inspired oxygen
* Units: not applicable
* Values: [0.21, 1.00]
  * expected floating point
  * 0.21 is room air
  * 1.00 is pure oxygen
* Caveats:
* Look back: 6 hours

### PFR: PaO~2~/FiO~2~
* Units: mmHg
* Values [0, Inf)
  * Expect floating point
* Caveats:
  * PFR is only valid if the time of FiO~2~ value is the same as, or before,
    the time of PaO~2~ value.  For example, if FiO~2~ was reported at 11:31 and
    PaO~2~ was reported at 11:45 then the PFR would be valid to use.  If the
    FiO~2~ is then updated at 11:59, the PaO~2~ value from 11:45 would not be
    valid to use in the calculation of the PFR

    | time  | FiO~2~           | PaO~2~          | PFR    | Note:                                                        |
    |   --: | ---:             | ---:            | ---:   | ------:                                                      |
    | 11:31 | 0.30*            |                 |        |                                                              |
    | 11:45 | 0.30<sup>c</sup> | 118*            | 393.33 |                                                              |
    | 11:47 | 0.30<sup>c</sup> | 118<sup>c</sup> | 393.33 |                                                              |
    | 11:59 | 0.40*            | 118<sup>c</sup> |        | No PFR, FiO~2~ is younger than PaO~2~                        |
    | 12:10 | 0.40<sup>c</sup> | 152*            | 380    |                                                              |
    | 12:15 | 0.40<sup>c</sup> | 152<sup>c</sup> | 380    |                                                              |
    | 17:45 | 0.40<sup>c</sup> | 152<sup>c</sup> | 380    |                                                              |
    | 18:00 |                  | 152<sup>c</sup> |        | No PFR, FiO~2~ carry forward window expired                  |
    | 18:30 |                  |                 |        | No PFR, FiO~2~ and PaO~2~ carry forward windows have expired |

      : * New value <br>
      <sup>c</sup> value carried forward

* Look back: implied by the look back on PaO~2~ and FiO~2~ and the condition
  that FiO~2~ value is the same age or older than the PaO~2~ value.

### SFR: SpO~2~ / FiO~2~
* Units: none
* Values: [0, 97/0.21]
  * Floating point
  * The upper limit includes the limitation on SpO~2~ ≤ 97
* Caveats:
  * SFR is only valid if the time of FiO~2~ value is the same as, or before,
    the time of SpO~2~ value.  For example, if FiO~2~ was reported at 11:31 and
    SpO~2~ was reported at 11:45 then the SFR would be valid to use.  If the
    FiO~2~ is then updated at 11:59, the SpO~2~ value from 11:45 would not be
    valid to use in the calculation of the SFR

     | time  | FiO~2~           | SpO~2~         | SFR   | Note:                                                        |
     | --:   | ---:             | ---:           | ---:  | ------:                                                      |
     | 10:15 | 0.25*            | 92*            | 368   |                                                              |
     | 10:20 | 0.25<sup>c</sup> | 98*            |       | No SFR, SpO~2~ > 97                                          |
     | 11:31 | 0.30*            | 98<sup>c</sup> |       | No SFR, SpO~2~ > 97 and FiO~2~ younger than SpO~2~           |
     | 11:45 | 0.30<sup>c</sup> | 87*            | 290   |                                                              |
     | 11:47 | 0.30<sup>c</sup> | 87<sup>c</sup> | 290   |                                                              |
     | 11:59 | 0.40*            | 87<sup>c</sup> |       | No SFR, FiO~2~ is younger than SpO~2~                        |
     | 12:10 | 0.40<sup>c</sup> | 85*            | 212.5 |                                                              |
     | 12:15 | 0.40<sup>c</sup> | 85<sup>c</sup> | 212.5 |                                                              |
     | 17:45 | 0.40<sup>c</sup> | 85<sup>c</sup> | 212.5 |                                                              |
     | 18:00 |                  | 85<sup>c</sup> |       | No SFR, FiO~2~ carry forward window expired                  |
     | 18:30 |                  |                |       | No SFR, FiO~2~ and SpO~2~ carry forward windows have expired |

      : * New value <br>
      <sup>c</sup> value carried forward

* Look back: implied by the look back on SpO~2~ and FiO~2~ and the condition
  that FiO~2~ value is the same age or older than the SpO~2~ value.

### ORS: Other Respiratory Support
* Units: not applicable
* Values:
  * 0: No other respiratory support
  * 1: some form of respiratory support
    * may or may not include supplemental oxygen
* Caveats:
  * A backstop: if FiO~2~ > 0.21 then that indicates oxygen support and thus
    ORS
* Look Back: 6 hours
  * If it has been more than six hours since last known data point indicating
    the patient has respiratory support then consider the patient to no longer
    have respiratory support.

### IMV: Invasive Mechanical ventilation
* Units: not applicable
* Values:
  * 0: patient is not currently intubated
  * 1: patient currently has invasive mechanical ventilation
* Caveats:
  * IMV due to surgery should not be used to assess organ dysfunction
* Look back: 6 hours
  * If it has been more than 6 hours since the last known data point
    indicating the patient has IMV, then consider the patient to no longer
    have IMV


## Scoring
Scoring is based on the worse case based on available PaO~2~/FiO~2~ (PFR) and
SpO~2~/FiO~2~ (SFR) with consideration for IMV and ORS.

```{r}
# R code: booleans are implicitly coerced to integers
cat(tail(as.character(body(phoenix::phoenix_respiratory)), 1), sep = "\n")
```

# Cardiovascular

Cardiovascular dysfunction is based on three conceptual sets:

1. Systemic Vasoactive mediations
2. Lactate
3. Age adjusted mean arterial pressure

## Inputs

### Systemic Vasoactive Medications

There are six systemic vasoactive medications to consider:
1. dobutamine
2. dopamine
3. epinephrine
4. norepinephrine
5. milrinone
6. vasopressin

Vasoactive medications (dobutamine, dopamine, etc.) should only count if they
are systemic. In our older dataset, epinephrine had some non-systemic medication
administrations. For example: "epinephrine hcl 2.25 % inh neb soln" should not
count towards the systemic vasoactive medication count towards cardiovascular
system dysfunction points.

The routes for administration of the vasoactive medications are at
Example SQL code used for mapping routes when developing Phoenix can be found in
a [GitHub archive](https://github.com/CU-DBMI-Peds/phoenix_sepsis_criteria/tree/main/harmonization/medication_mappings)
specifically the files:
[med_route_mapping.sql](https://github.com/CU-DBMI-Peds/phoenix_sepsis_criteria/blob/main/harmonization/medication_mappings/med_route_mapping.sql)
and
[clean_up_routes.sql](https://github.com/CU-DBMI-Peds/phoenix_sepsis_criteria/blob/main/harmonization/medication_mappings/clean_up_routes.sql).

The overall meta data for vasoactives:

* Units: none
* Values: 0, 1, 2, 3, 4, 5, 6
* Caveats: see above
* Look back: 12 hours


### Lactate
* Units: mmol/L
* Values: [0, 50]
* Caveats:
* Look back: 6 hours

### Age
* Units: Months
* Values: [0, 216)
* Caveats:
* Look back:

### Mean Arterial Pressure (MAP)

There can be many sources of blood pressure.  We used the following hierarchy for
which value to use in the scoring.

1. Arterial MAP (direct measurement)
2. Estimated MAP from arterial systolic (SBP) and arterial diastolic (DBP) pressures
3. Cuff MAP (from a machine)
4. Estimated MAP from cuff SBP and cuff DBP

MAP was estimated as (2/3) * DBP + (1/3) * SBP

Expected metadata:

* MAP:
  * Units: mmHg
  * Values: [0, 300]
  * Caveats:
    * If MAP is provided along with SBP and DBP it is worth verifying that DBP ≤ MAP ≤ SBP
  * Look back: 6 hours

* SBP:
  * Units: mmHg
  * Values: [0, 300]
  * Caveats:
    * It is worth verifying that DBP ≤ SBP, we used a slightly more
      conservative check of (SBP - DBP) ≥ 1
  * Look back: 6 hours

* DBP:
  * Units: mmHg
  * Values: [0, 200]
  * Caveats:
    * It is worth verifying that DBP ≤ SBP, we used a slightly more
      conservative check of (SBP - DBP) ≥ 1
  * Look back: 6 hours

## Scoring

Up to six points can be earned from cardiovascular dysfunction.  Up to two
points for vasoactives, up to two points for lactate, and up to two points for
MAP.

```{r}
# vas: number of vasoactives meds
# lct: lactate in mmol/L
# age: age in months
# map: mean arterial pressure in mmHg
cat(tail(as.character(body(phoenix::phoenix_cardiovascular)), 4), sep = "\n")
```

# Coagulation

## Platelets
## INR
## D-dimer
## Fibrinogen

# Neurologic

## Inputs

### Glasgow Coma Scale (GCS)

Phoenix is based on the total GCS score
* GCS (total):
  * Units: none
  * Values: integer values 3, 4, 5, ..., 15
  * Caveats:
    * The total GCS is the sum of three component scores, eye, motor, and verable:
    * We assumed that if we had an updated score for any one of the three componets
      but not all, then the non-updated componets were assumed to have not
      changed and did not warrant data entry.  We, therefore, updated the time
      of the GCS total and all its compoents to the youngest update time for any
      of the compoents.
  * Look back: 12 hours

* GCS (eye):
  * Units: none
  * Values: integer values 1, 2, 3, 4
  * Caveats:
  * Look back: 12 hours

* GCS (motor):
  * Units: none
  * Values: integer values 1, 2, 3, 4, 5, 6
  * Caveats:
  * Look back: 12 hours

* GCS (verbal):
  * Units: none
  * Values: integer values 1, 2, 3, 4, 5
  * Caveats:
  * Look back: 12 hours

### Pupils
* Units: none
* Values:
  * 1: bilaterally fixed pupils
  * 0: one or zero fixed pupils
* Caveats:
* Look back: 6 hours

## Scoring

* At most 2 points.
* 1 point for GCS ≤ 10
* 2 points for bilaterally fixed pupils

A patient with a GCS ≤ 10 and bilaterally fixed pupils will have a score of 2.

```{r}
# fpl: 1 if bilaterally fixed pupils; 0 otherwise
cat(tail(as.character(body(phoenix::phoenix_neurologic)), 1), sep = "\n")
```

# Endocrine

## Inputs
### Glucose
* Units: mg/dL
* Values: [0, Inf)
  * floating point
* Caveats:
* Look back: 12 hours

## Scoring
Scoring for Phoenix-8 only.  1 point if glucose < 50 or > 150

```{r}
cat(tail(as.character(body(phoenix::phoenix_endocrine)), 1), sep = "\n")
```

# Immunologic
## ANC
## ALC

# Renal
## Age (see notes in cardiovascular)
## Creatinine

# Hepatic

## Total Bilirubin

Units should be in mg/dL, or converted to mg/dL and range from 0 - 100.

## ALT

# References
