<!-- README.md is generated from README.Rmd. Please edit that file -->

# phoenix: Phoenix Sepsis and Phoenix-8 Sepsis Criteria

Implementation of the Phoenix and Phoenix-8 Sepsis Criteria as
described in:

* ["Development and Validation of the Phoenix Criteria for Pediatric Sepsis and Septic Shock"](doi:10.1001/jama.2024.0196) by Sanchez-Pinto&ast;, Bennett&ast;, DeWitt&ast;&ast;, Russell&ast;&ast; et al. (2024)

  * <small> &ast; Drs Sanchez-Pinto and Bennett contributed equally; &ast;&ast; Dr DeWitt and Mr Russell contributed equally.</small>

* ["International Consensus Criteria for Pediatric Sepsis and Septic Shock"](doi:10.1001/jama.2024.0179) by Schlapbach, Watson, Sorce, et al. (2024).

## Phoenix Criteria
A patient is consided septic if they have a suspected (or proven) infection with
at total Phoenix score of at least two points.  The Phoenix score is the sum of
the scores from four organ dysfunction scores:

1. Respiratory,
2. Cardiovascular,
3. Coagulation, and
4. Neurologic.

Septic shock is defined as sepsis with at least one cardiovascular point.

In addition to the Phoenix criteria there is an extended crieria intended for
research, Phoenix-8, which includes the four organ systems above along with

5. Endocrine,
6. Immunologic,
7. Renal, and
8. Hepatic.

## Phoenix Rubric

```{r, echo = FALSE, results = "asis"}
# Need to use github flavored markdown for this readme which means chaning the
# pandoc flavor to github flavor
tab <-
  scan(file = system.file("phoenix_rubric.md", package = "phoenix"),
       what = character(),
       sep = "\n",
       quiet = TRUE)
tab <- gsub("\\^(\\w)\\^", "<sup>\\1</sup>", tab)
tab <- gsub("\\~(\\w)\\~", "<sub>\\1</sub>", tab)
cat(tab, sep = "\n")
cat("\n")
```

<small>
```{r, echo = FALSE, results = "asis"}
# Need to use github flavored markdown for this readme which means chaning the
# pandoc flavor to github flavor
ftnt <-
  scan(file = system.file("phoenix_rubric_footnotes.md", package = "phoenix"),
       what = character(),
       sep = "\n",
       quiet = TRUE)
ftnt <- gsub("\\^(\\w)\\^", "<sup>\\1</sup>", ftnt)
ftnt <- gsub("\\~(\\w)\\~", "<sub>\\1</sub>", ftnt)
cat(ftnt[1:9], sep = "\n\n")
```
</small>


## Software

* Python module
* R package: [![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/phoenix)](https://cran.r-project.org/package=phoenix)
* [Example SQL queries](https://cu-dbmi-peds.github.io/phoenix/articles/sql.html)

## Example use of the Python Module

```{r, label = setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
library(reticulate)

if (!virtualenv_exists(envname = "phoenix")) {
  virtualenv_create(envname = "phoenix")
  virtualenv_install(envname = "phoenix", packages = c("numpy", "pandas"))
  virtualenv_install(envname = "phoenix", packages = normalizePath('.'))
}

use_virtualenv(virtualenv = "phoenix")
```

```{python}
import numpy as np
import pandas as pd
import importlib.resources

import phoenix as phx
```

There is an example data set included in the package in a file called
`sepsis.csv`.  Load that file to use in the following examples.

```{python}
# read in the example data set
path = importlib.resources.files('phoenix')
sepsis = pd.read_csv(path.joinpath('data').joinpath('sepsis.csv'))
print(sepsis.head())
```

### Organ Dysfunction Scoring
There is one funciton for each of the eight component organ dysfunction scores.

#### Respiratory Dysfunction Scoring

```{python}
resp = phx.phoenix_respiratory(
    pf_ratio = sepsis["pao2"] / sepsis["fio2"],
    sf_ratio = np.where(sepsis["spo2"] <= 97, sepsis["spo2"] / sepsis["fio2"], np.nan),
    imv      = sepsis["vent"],
    other_respiratory_support = (sepsis["fio2"] > 0.21).astype(int).to_numpy()
    )
print(type(resp))
print(resp)
```

#### Cardiovascular Dysfunction Scoring

```{python}
card = phx.phoenix_cardiovascular(
    vasoactives = sepsis["dobutamine"] + sepsis["dopamine"] + sepsis["epinephrine"] + sepsis["milrinone"] + sepsis["norepinephrine"] + sepsis["vasopressin"],
    lactate = sepsis["lactate"],
    age = sepsis["age"],
    map = sepsis["dbp"] + (sepsis["sbp"] - sepsis["dbp"]) / 3
)
print(type(card))
print(card)
```

#### Coagulation Dysfunction Scoring

```{python}
coag = phx.phoenix_coagulation(
    platelets = sepsis['platelets'],
    inr = sepsis['inr'],
    d_dimer = sepsis['d_dimer'],
    fibrinogen = sepsis['fibrinogen']
)
print(type(coag))
print(coag)
```

#### Neurologic Dysfunction Scoring

```{python}
neuro = phx.phoenix_neurologic(
    gcs = sepsis["gcs_total"],
    fixed_pupils = (sepsis["pupil"] == "both-fixed").astype(int)
)
print(type(neuro))
print(neuro)
```

#### Endocrine Dysfunction Scoring

```{python}
endo = phx.phoenix_endocrine(sepsis["glucose"])
print(type(endo))
print(endo)
```

#### Immunologic Dysfunction Scoring

```{python}
immu = phx.phoenix_immunologic(sepsis["anc"], sepsis["alc"])
print(type(immu))
print(immu)
```

#### Renal Dysfunction Scoring

```{python}
renal = phx.phoenix_renal(sepsis["creatinine"], sepsis["age"])
print(type(renal))
print(renal)
```

#### Hepatic Dysfunction Scoring

```{python}
hepatic = phx.phoenix_hepatic(sepsis["bilirubin"], sepsis["alt"])
print(type(hepatic))
print(hepatic)
```

### Phoenix Scoring

```{python}
print("Phoenix Sepsis Score")
phoenix = phx.phoenix(
    pf_ratio = sepsis["pao2"] / sepsis["fio2"],
    sf_ratio = sepsis["spo2"] / sepsis["fio2"],
    imv      = sepsis["vent"],
    other_respiratory_support = (sepsis["fio2"] > 0.21).astype(int).to_numpy(),
    vasoactives = sepsis["dobutamine"] + sepsis["dopamine"] + sepsis["epinephrine"] + sepsis["milrinone"] + sepsis["norepinephrine"] + sepsis["vasopressin"],
    lactate = sepsis["lactate"],
    age = sepsis["age"],
    map = sepsis["dbp"] + (sepsis["sbp"] - sepsis["dbp"]) / 3,
    platelets = sepsis['platelets'],
    inr = sepsis['inr'],
    d_dimer = sepsis['d_dimer'],
    fibrinogen = sepsis['fibrinogen'],
    gcs = sepsis["gcs_total"],
    fixed_pupils = (sepsis["pupil"] == "both-fixed").astype(int),
    )

print(type(phoenix))
print(phoenix.head())
```

### Phoenix-8 Scoring

```{python}
phoenix8_scores = phx.phoenix8(
    pf_ratio = sepsis["pao2"] / sepsis["fio2"],
    sf_ratio = np.where(sepsis["spo2"] <= 97, sepsis["spo2"] / sepsis["fio2"], np.nan),
    imv      = sepsis["vent"],
    other_respiratory_support = (sepsis["fio2"] > 0.21).astype(int).to_numpy(),
    vasoactives = sepsis["dobutamine"] + sepsis["dopamine"] + sepsis["epinephrine"] + sepsis["milrinone"] + sepsis["norepinephrine"] + sepsis["vasopressin"],
    lactate = sepsis["lactate"],
    map = sepsis["dbp"] + (sepsis["sbp"] - sepsis["dbp"]) / 3,
    platelets = sepsis['platelets'],
    inr = sepsis['inr'],
    d_dimer = sepsis['d_dimer'],
    fibrinogen = sepsis['fibrinogen'],
    gcs = sepsis["gcs_total"],
    fixed_pupils = (sepsis["pupil"] == "both-fixed").astype(int),
    glucose = sepsis["glucose"],
    anc = sepsis["anc"],
    alc = sepsis["alc"],
    creatinine = sepsis["creatinine"],
    bilirubin = sepsis["bilirubin"],
    alt = sepsis["alt"],
    age = sepsis["age"]
    )

print(type(phoenix8_scores))
print(phoenix8_scores.head())
```

